sem

### 2023-01-31 scripts/workshop/sem.Rmd
add text and knit to html

Make lagged data - I'll use this later to compare results using lagged and not lagged data
```{r}
prev_cc <- comm_data

@@

#remove the first entry for each wetland to remove the carryover from the last wetland
clean <- order %>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)

@@

In this first iteration, I first propose the relationships independently and then combine them later to generate the SEM. Each relationship is modeled as a linear regression and is evaluated separately. The 'piecewiseSEM' package generates a test statistic (Fisher's C) and and AIC score through a method known as "tests of directed seperation". This method evalutes the relationships between variables that are *un*specified in the model and asks whether these relationships are truly independent or whether some causal relationship may exist. piecewiseSEM relies on a priori knowledge of the system that can be used to rationalize expected relationships in the data. The benefits of piecewiseSEM is that it allows for non-independence of the data and allows for each relationship to be evaluated independently using

cc ~ mean water temp; 
size ~ mean water temp; 
prevalence ~ cc + size + mean water temp

@@

```{r}
comm_data %<>% filter(MeanWaterTempPredC>0)

cc_temp <- gam(log10(cc) ~ s(MeanWaterTempPredC), data = comm_data)
size_temp <- lm(log10(size) ~ (MeanWaterTempPredC), data = comm_data)
prev_cc <- lm(Prevalence ~ log10(cc) + log10(size), data = comm_data)
prev_temp_cc <- glm(Prevalence ~ log10(cc) + log10(size) + MeanWaterTempPredC, data = comm_data)

@@

#fix error
#sem <- psem(cc_temp, size_temp, prev_cc)
#sem_1 <- psem(cc_temp, size_temp, prev_temp_cc)
```

Using lagged data
```{r}
lagged_cc_temp <- lm(lag_cc ~ lag_temp, data = clean)
lagged_size_temp <- lm(log10(lag_size) ~ lag_temp + lag_cc, data = clean)
lagged_prev_cc <- lm(Prevalence ~ lag_cc + log10(lag_size), data = clean)
lagged_prev_temp_cc <- lm(Prevalence ~ lag_cc + log10(lag_size) + lag_temp, data = clean)

@@

@@

@@

@@

>>>>

Make lagged data - I'll use this later to compare results using lagged and not lagged data

```{r}
prev_cc <- comm_data

@@

#remove the first entry for each wetland to remove the carryover from the last wetland
clean <- order %>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)
```

```{r}
comm_data %<>% filter(!is.na(MeanWaterTempPredC)) #remove missing water temp data
clean %<>% filter(!is.na(MeanWaterTempPredC)) 
```

@@

In this first iteration, I first propose the relationships independently and then combine them later to generate the SEM. Each relationship is modeled as a linear regression and is evaluated separately. The 'piecewiseSEM' package generates a test statistic (Fisher's C) and and AIC score through a method known as "tests of directed seperation". This method evalutes the relationships between variables that are *un*specified in the model and asks whether these relationships are truly independent or whether some causal relationship may exist. piecewiseSEM relies on a priori knowledge of the system that can be used to rationalize expected relationships in the data. The benefits of piecewiseSEM is that it allows for non-independence of the data and allows for each relationship to be evaluated independently using a more flexible range of modeling options (e.g. GLM and GAM). I develop two sets of SEMs here. One assumes prevalence is only influenced directly by cc and size and the other assumes that water temp may also have a direct effect on prevalence.

cc ~ mean water temp; 
size ~ mean water temp; 
prevalence ~ cc
prevalence ~ cc + size + mean water temp

@@

I attempt to use a gam to model the relationship between water temp and cc and a glm to model the effects of cc, size, and temp on infection prevalence but I keep getting errors and can't figure out how to fix it. For some reason, I am losing all of the data during this process.

```{r}
cc_temp <- gam(log10(cc) ~ s(MeanWaterTempPredC), data = comm_data)
size_temp <- lm(log10(size) ~ (MeanWaterTempPredC), data = comm_data)
prev_cc <- lm(Prevalence ~ log10(cc) + log10(size), data = comm_data)
prev_temp_cc <- glm(Prevalence ~ log10(cc) + log10(size) + MeanWaterTempPredC, data = comm_data)

@@

#fix error
#Error in data.frame(..., check.names = FALSE) :
#arguments imply differing number of rows: 0, 83
# sem <- psem(cc_temp, size_temp, prev_cc)
# sem_1 <- psem(cc_temp, size_temp, prev_temp_cc)
```

Same relationships using lagged data

```{r}
lagged_cc_temp <- lm(lag_cc ~ lag_temp, data = clean)
lagged_size_temp <- lm(log10(lag_size) ~ lag_temp + lag_cc, data = clean)
lagged_prev_cc <- lm(Prevalence ~ lag_cc + log10(lag_size), data = clean)
lagged_prev_temp_cc <- lm(Prevalence ~ lag_cc + log10(lag_size) + lag_temp, data = clean)

@@

The following chunks goes through the piecewiseSEM steps individually.

@@

The following sections shows the same information as before but using summary() instead of doing it individually. I also plot all of the SEMs that were previously evaluated.

@@

In the unlagged data, there are no significant relationships between any variables and prevalence. Also, the model that does not include a direct relationship between temperature and prevalence has a slightly lower (35.971 vs. 36.545) AIC than the other.

@@

In the lagged data, there is no significant direct relationship between temperature and prevalence but there is a signficant relationship between temperature and cc and cc and prevalence. This would mean that there is an indirect relationship between temperature and prevalence that is mediated by cc. Temperature has a negative effect on cc but cc has a positive effect on prevalence.
